---
title: "Prepping IUCN and AM data for file size"
format: 
  html:
    code-fold: true
    code-summary: "Show me the code"
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(terra)
library(jsonlite)
library(here)
```

# Prep data for live coding


## Prep IUCN data

This set of species includes all marine-associated mammals only. For now, note IUCN species ID, scientific and common name, and risk category, among other information. The risk category is what we want - up to date as of the end of 2021.

```{r}

iucn_dir <- '/home/shares/ohi/spp_vuln/spp_vuln_mapping/iucn_spp'
full_data_f <- file.path(iucn_dir, 'spp_info_from_api_2021-3.csv')
full_df <- read_csv(full_data_f)

hab_data_f <- file.path(iucn_dir, 'spp_habs_from_api_2021-3.csv')
hab_df <- read_csv(hab_data_f)

hab_cols <- names(hab_df)

mar_mamm_df <- full_df %>%
  filter(class == 'MAMMALIA') %>%
  left_join(hab_df, by = 'iucn_sid') %>%
  filter(str_detect(habitat, '^Marine')) %>%
  filter(suitability == 'Suitable') %>%
  select(everything(), -all_of(hab_cols), iucn_sid) %>%
  distinct()
write_csv(mar_mamm_df, here('data/iucn_mar_mammal_info_v2021-3.csv'))
```

## Prep AquaMaps species info

This set of species includes all marine mammals only. For now, note AquaMaps species ID, IUCN species ID, name, sciname, and risk category, among other information. Here, the risk category is not up to date - this dataset was last updated in 2019.

```{r}
am_dir <- '/home/shares/ohi/git-annex/aquamaps_2021'
am_spp_f <- file.path(am_dir, 'ver10_2019_speciesoccursum_iucn.csv')
am_spp_df <- read_tsv(am_spp_f) %>% ### why is this tab separated?
  filter(Class == 'Mammalia')
write_csv(am_spp_df, here('data/am_mammal_info_v10_2019.csv'))
```

```{r}
base_r <- rast(xmin = -180, xmax = 180, ymin = -90, ymax = 90, resolution = 0.5, crs = 'epsg:4326') %>%
  setValues(1:ncell(.))
ca_ext <- ext(c(-150, -100, 0, 65))

cropped_r <- base_r %>%
  crop(ca_ext)
loiczid_vec <- values(cropped_r)

hcaf_f <- file.path(am_dir, 'hcaf_v7.csv')
hcaf_cropped_df <- read_csv(hcaf_f) %>%
  filter(LOICZID %in% loiczid_vec)

write_csv(hcaf_cropped_df, here('data/am_hcaf_cropped_v7.csv'))

am_spp_cell_f <- file.path(am_dir, 'hcaf_species_native_clean.csv')
am_spp_cell_df <- data.table::fread(am_spp_cell_f)
am_spp_cell_mamm <- am_spp_cell_df %>%
  filter(am_sid %in% am_spp_df$SpeciesID) %>%
  filter(loiczid %in% loiczid_vec)

write_csv(am_spp_cell_mamm, here('data/am_mammal_hcaf_cropped_v10_2019.csv'))

```

