---
title: "Exploring some ecological datasets"
format: 
  html:
    code-fold: true
    code-summary: "Show me the code"
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(terra) ### remotes::install_github('rspatial/terra')
library(sf)
library(jsonlite)
library(here)
library(rnaturalearth)
```

# Overview

## IUCN API

The IUCN API documentation is here: https://apiv3.iucnredlist.org/api/v3/docs

All API calls will start with most of that as a stem. We will change the endpoints to access different types of information, sometimes requiring us to include additional parameters.

For most endpoints we will also need an authorization token.

Try one of the examples, and note how the URL is filled in with different parameters and token=XXXXX. We'll borrow that token for our work today, but if you do real work with this API, request your own token.

Use `usethis::edit_r_environ()` to open `.Renviron` file. Type a key name, then paste in key, then save and close. Access using `Sys.getenv()`

```{r}
iucn_token <- Sys.getenv('IUCN_KEY')
```

### Access info on many species (one page)

Assemble a URL and pass it parameters. Inspect the JSON (JavaScript Object Notation). Read it in as a dataframe using `jsonlite::fromJSON()`

```{r}
iucn_url <- 'https://apiv3.iucnredlist.org'
spp_page_endpoint <- paste0(iucn_url, '/api/v3/species/page/%s?token=%s')

spp_page_url <- sprintf(spp_page_endpoint, 2, iucn_token)

spp_page <- jsonlite::fromJSON(spp_page_url)

spp_page_df <- spp_page$result # or spp_page[['result']]

```

### Access specific info on a single species

Humpback whale is ID \# 13006. Let's find the countries where it can be found! (or try a different taxon ID from the previous step)

```{r}
spp_country_endpoint <- paste0(iucn_url, '/api/v3/species/countries/id/%s?token=%s')
spp_country_url <- sprintf(spp_country_endpoint, 13006, iucn_token)

spp_ctry <- jsonlite::fromJSON(spp_country_url)
spp_ctry_df <- spp_ctry$result

DT::datatable(spp_ctry_df)
```

## WoRMS API

Now let's try it with the World Register of Marine Species! https://www.marinespecies.org/rest

### Access AphiaID for a genus name

Let's find the ID for genus Balaenoptera, a group that includes the biggest whales in the oceans.

```{r}
worms_url <- 'https://www.marinespecies.org/rest/'
worms_id_endpoint <- paste0(worms_url, 'AphiaIDByName/%s')
worms_id_url <- sprintf(worms_id_endpoint, 'balaenoptera')
worms_id <- jsonlite::fromJSON(worms_id_url)

```

### Access children of a genus

The ID for Balaenoptera is 137013, but it is stored in a variable already. Let's find the "children", i.e., the members of the next classification down (genus -\> species)

```{r}
worms_child_endpoint <- paste0(worms_url, 'AphiaChildrenByAphiaID/%s')
worms_child_url <- sprintf(worms_child_endpoint, worms_id)
worms_child <- jsonlite::fromJSON(worms_child_url)

worms_child_acc <- worms_child %>%
  filter(status == 'accepted')

DT::datatable(worms_child)
```

### Access classification of a species

Not all returned data are equally easy to work with - here's a list of lists of lists.

```{r}
worms_class_endpoint <- paste0(worms_url, 'AphiaClassificationByAphiaID/%s')
worms_class_url <- sprintf(worms_class_endpoint, worms_id)
worms_class <- jsonlite::fromJSON(worms_class_url)

```

Many coders have developed packages to work with specific APIs - the ROpenSci `taxize` package works well with taxonomic info from WoRMS and several other sources. There's also a `rredlist` package for working with the IUCN RedList API.

## Read in data

### read in cleaned IUCN data

This set of species includes all marine-associated mammals only.  For now, note IUCN species ID, scientific and common name, and risk category, among other information.  The risk category is what we want - up to date as of the end of 2021.

```{r}
iucn_spp_df <- read_csv(here('data/iucn_mar_mammal_info_v2021-3.csv'))
```

### Read in cleaned AquaMaps species info

This set of species includes all marine mammals only.  For now, note AquaMaps species ID, IUCN species ID, name, sciname, and risk category, among other information.  Here, the risk category is not up to date - this dataset was last updated in 2020.

The critical piece here is the AquaMaps species ID, which we need to know for the spatial information in a moment.  We'll use IUCN id to join the up-to-date risk category to the AquaMaps species ID.

```{r}
am_spp_df <- read_csv(here('data/am_mammal_info_v10_2019.csv')) %>%
  janitor::clean_names() %>%
  rename(am_sid = species_id, iucn_sid = iucn_id)
```

### Read in cleaned AquaMaps species-cell info

Note the columns loiczid, AquaMaps species id, and probability of occurrence.  LOICZID is a cell code that we'll match to spatial information in a minute.

```{r}
am_spp_cells_df <- read_csv(here('data/am_mammal_hcaf_cropped_v10_2019.csv'))
```


### Read in cleaned HCAF data

This is cropped to the Eastern Pacific Ocean - east of -150 longitude, north of the equator.  We will turn this into a raster for mapping our species!

```{r}
am_hcaf_df <- read_csv(here('data/am_hcaf_cropped_v7.csv')) %>%
  janitor::clean_names()

longlat_id_df <- am_hcaf_df %>%
  filter(ocean_area > 0) %>%
  select(center_long, center_lat, loiczid)

hcaf_r <- rast(longlat_id_df, type = 'xyz', crs = 'epsg:4326')

plot(hcaf_r)
```

## Start joining

We want to map IUCN risk codes against AquaMaps spatial data.  Join the IUCN and AquaMaps information, to get the risk codes attached to the AquaMaps ids.  

Then, using AquaMaps IDs, join the risk codes to the LOICZID information.  Some mammal spp don't occur in our selected region (e.g., South African fur seal), so end up with NA for LOICZID.

```{r}
id_risk_df <- iucn_spp_df %>%
  select(iucn_sid, sciname, category) %>%
  mutate(cat_score = case_when(category == 'LC' ~ 0,
                               category == 'NT' ~ 1,
                               category == 'VU' ~ 2,
                               category == 'EN' ~ 3,
                               category == 'CR' ~ 4,
                               category == 'EX' ~ 5,
                               TRUE ~ NA_real_)) %>%
  left_join(am_spp_df, by = 'iucn_sid') %>%
  select(iucn_sid, am_sid, sciname, f_bname, category, cat_score) %>%
  filter(!is.na(am_sid))

risk_cells_df <- id_risk_df %>%
  left_join(am_spp_cells_df, by = 'am_sid')
```

## Analysis 1: map a single species by relative environmental suitability

Let's look at suitability of environments for killer whales, aka _Orcinus orca_.  We'll create a raster based on our cropped HCAF, by substituting in values of prob for values of LOICZID.
```{r}
orca_df <- risk_cells_df %>%
  filter(sciname == 'Orcinus orca')

orca_prob_r <- terra::subst(hcaf_r, 
                            from = orca_df$loiczid, 
                            to = orca_df$prob,
                            others = NA) ### "others" is in dev version
  
plot(orca_prob_r,
     main = 'probability of occurrence for Orcinus orca')
```

### Let's ggplot it

``` {r}
orca_prob_df <- as.data.frame(orca_prob_r, xy = TRUE) %>%
  rename(prob = loiczid)

land_sf <- ne_countries(scale = 110, 
                        continent = 'north america',
                        returnclass = 'sf')

ggplot() +
  geom_raster(data = orca_prob_df, aes(x, y, fill = prob)) +
  scale_fill_viridis_c() +
  geom_sf(data = land_sf) +
  coord_sf(xlim = c(-150, -100), ylim = c(0, 60)) +
  theme_void()

```


## Analysis 2: species richness

Here we can group by cell and count up the number of spp found.  Note, we may want to not count species with low probability of occurrence - let's apply a filter for probability at least 50%.

```{r}
spp_rich_df <- risk_cells_df %>%
  filter(prob >= 0.5) %>%
  group_by(loiczid) %>%
  summarize(n_spp = n_distinct(am_sid))

spp_rich_r <- terra::subst(hcaf_r, 
                           from = spp_rich_df$loiczid, 
                           to = spp_rich_df$n_spp,
                           others = NA)
  
plot(spp_rich_r,
     main = 'Species richness of marine mammals')
```

### ggplot at your leisure

## Analysis 3: mean risk

Here we can group by cell and find the average risk.  First, let's see how many spp in each risk category.

```{r}
table(id_risk_df$category) ### note a few DD

mean_risk_df <- risk_cells_df %>%
  filter(prob >= 0.5) %>%
  group_by(loiczid) %>%
  summarize(mean_risk = mean(cat_score))
  # summarize(mean_risk = mean(cat_score, na.rm = TRUE))

mean_risk_r <- terra::subst(hcaf_r, 
                           from = mean_risk_df$loiczid, 
                           to = mean_risk_df$mean_risk,
                           others = NA)
  
plot(mean_risk_r,
     main = 'Species richness of marine mammals')
```


